---
title: "Forecasting Dengue in Sri Lanka: A Journey Through Challenges, Actions, and Lessons Learned"
format:
  revealjs: 
    theme: [default]
    width: 1280
    height: 720
    slide-number: true
    title-slide-attributes: 
      data-background-color: "#3f007d"
      data-background-image: none
author:
  - name: Thiyanga S. Talagala
    affiliations: Department of Statistics, University of Sri Jayewardenepura, Sri Lanka 
subtitle: "Women in Forecasting (WIF) Conference - 2026, Montreal, Canada"
---

## {background-image="map1.png" background-size="cover"}

## {background-image="map2.png" background-size="cover"}

## {background-image="map3.png" background-size="cover"}

## Sri Lanka climate 

🌴Climate type: 

- Tropical climate: Warm and humid throughout the year

 ☀️Average temperature:
 
  - Typically 26°C – 30°C in lowlands
    
  - Cooler in central highlands (e.g., Nuwara Eliya ~16°C–20°C)

 🌧💧️ Rainfall:


  - Two main monsoon seasons: 
  
    - Southwest monsoon (May–September)
    
    - Northeast monsoon (December–February)
    
    
 
# {background-image="map4.jpg" background-size="contain"} 

# Challenges




# {background-image="map5.png" background-size="contain" background-repeat="no-repeat" background-position="center"}


#  {background-image="map6.png" background-size="contain" background-repeat="no-repeat" background-position="center"}

# What makes dengue forecasting a challenging problem?



## Weekly Dengue Cases in Sri Lanka: 2007 - 2016


```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(denguedatahub)
library(dplyr)
library(ggplot2)
library(lubridate)

srilanka_weekly_data %>%
  mutate(
    start.date = mdy(start.date),
    year = year(start.date)
  ) %>%
  group_by(start.date, year) %>%
  summarise(island_cases = sum(cases, na.rm = TRUE), .groups = "drop") %>%
  mutate(period = ifelse(year <= 2016, "≤ 2017", "> 2017")) %>%
  ggplot(aes(x = start.date, y = island_cases, color = period)) +
  geom_line() +
  scale_color_manual(values = c("≤ 2017" = "black", "> 2017" = "white")) +
  #theme_bw() +
  labs(
   # title = "Weekly Dengue Cases in Sri Lanka (Island Total)",
    x = "Year",
    y = "Cases",
    color = "Period"
  ) + theme(legend.position = "none")
```


## Weekly Dengue Cases in Sri Lanka: 2007 - 2018


```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(denguedatahub)
library(dplyr)
library(ggplot2)
library(lubridate)

srilanka_weekly_data %>%
  mutate(
    start.date = mdy(start.date),
    year = year(start.date)
  ) %>%
  group_by(start.date, year) %>%
  summarise(island_cases = sum(cases, na.rm = TRUE), .groups = "drop") %>%
  mutate(period = ifelse(year <= 2018, "≤ 2017", "> 2017")) %>%
  ggplot(aes(x = start.date, y = island_cases, color = period)) +
  geom_line() +
  scale_color_manual(values = c("≤ 2017" = "black", "> 2017" = "white")) +
  #theme_bw() +
  labs(
   # title = "Weekly Dengue Cases in Sri Lanka (Island Total)",
    x = "Year",
    y = "Cases",
    color = "Period"
  ) + theme(legend.position = "none")
```

## Weekly Dengue Cases in Sri Lanka: 2007 - 2019


```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(denguedatahub)
library(dplyr)
library(ggplot2)
library(lubridate)

srilanka_weekly_data %>%
  mutate(
    start.date = mdy(start.date),
    year = year(start.date)
  ) %>%
  group_by(start.date, year) %>%
  summarise(island_cases = sum(cases, na.rm = TRUE), .groups = "drop") %>%
  mutate(period = ifelse(year <= 2020, "≤ 2017", "> 2017")) %>%
  ggplot(aes(x = start.date, y = island_cases, color = period)) +
  geom_line() +
  scale_color_manual(values = c("≤ 2017" = "black", "> 2017" = "white")) +
  #theme_bw() +
  labs(
   # title = "Weekly Dengue Cases in Sri Lanka (Island Total)",
    x = "Year",
    y = "Cases",
    color = "Period"
  ) + theme(legend.position = "none")
```

## Weekly Dengue Cases in Sri Lanka: 2007 - 2022


```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(denguedatahub)
library(dplyr)
library(ggplot2)
library(lubridate)

srilanka_weekly_data %>%
  mutate(
    start.date = mdy(start.date),
    year = year(start.date)
  ) %>%
  group_by(start.date, year) %>%
  summarise(island_cases = sum(cases, na.rm = TRUE), .groups = "drop") %>%
  mutate(period = ifelse(year <= 2022, "≤ 2017", "> 2017")) %>%
  ggplot(aes(x = start.date, y = island_cases, color = period)) +
  geom_line() +
  scale_color_manual(values = c("≤ 2017" = "black", "> 2017" = "white")) +
  #theme_bw() +
  labs(
   # title = "Weekly Dengue Cases in Sri Lanka (Island Total)",
    x = "Year",
    y = "Cases",
    color = "Period"
  ) + theme(legend.position = "none")
```



## Dengue Cases in Sri Lanka: 2007 - 2026 March


```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(denguedatahub)
library(dplyr)
library(ggplot2)
library(lubridate)

srilanka_weekly_data %>%
  mutate(
    start.date = mdy(start.date),
    year = year(start.date)
  ) %>%
  group_by(start.date, year) %>%
  summarise(island_cases = sum(cases, na.rm = TRUE), .groups = "drop") %>%
  mutate(period = ifelse(year <= 2027, "≤ 2017", "> 2017")) %>%
  ggplot(aes(x = start.date, y = island_cases, color = period)) +
  geom_line() +
  scale_color_manual(values = c("≤ 2017" = "black", "> 2017" = "white")) +
  #theme_bw() +
  labs(
   # title = "Weekly Dengue Cases in Sri Lanka (Island Total)",
    x = "Year",
    y = "Cases",
    color = "Period"
  ) + theme(legend.position = "none")
```

# Historical dengue time series patterns may not persist due to changes in transmission dynamics.

# 

### Weekly Dengue Cases by District (2020 onwards)

```{r}
#| fig.width: 15
#| fig.height: 8
library(dplyr)
library(ggplot2)
library(lubridate)
library(viridis)
library(scales)

srilanka_weekly_data %>%
  mutate(start.date = mdy(start.date)) %>%
  filter(start.date >= as.Date("2020-01-01")) %>%
  group_by(district, start.date) %>%
  summarise(cases = sum(cases, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = start.date, y = district, fill = cases)) +
  geom_tile() +
  scale_fill_viridis_c(
    option = "C",
    limits = c(1, NA),
    breaks = c(1, 10, 50, 100, 200),   # marks on scale
    oob = scales::squish
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  ) +
  labs(
    x = "Time",
    y = "District",
    fill = "Cases"
  )
```


# Dengue is endemic in all districts across the country throughout the year.




#

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)

srilanka_weekly_data %>%
  mutate(start.date = mdy(start.date)) %>%
  filter(district %in% c("Colombo", "Gampaha", "Kalutara")) %>%
  arrange(start.date) %>%
  ggplot(aes(x = start.date, y = cases)) +
  geom_line(color = "steelblue", linewidth = 0.6) +
  facet_grid(rows = vars(district), scales = "free_y") +
  labs(
    title = "Weekly Dengue Cases: Colombo, Gampaha, Kalutara",
    x = "Week",
    y = "Number of Cases"
  ) +
 # theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

#

:::: {.columns}

::: {.column width="50%"}

```{r}
#| fig.width: 5
#| fig.height: 7
library(sf)
library(dplyr)
library(ggplot2)
library(SriLanka)

data("lka_adm2")

west_districts <- lka_adm2 %>%
  filter(NAME %in% c("Colombo District",
                     "Gampaha District",
                     "Kalutara District"))

ggplot() +
  geom_sf(data = lka_adm2, color = "black", linewidth = 0.2) +
  geom_sf(data = west_districts, fill = "steelblue", color = "black", linewidth = 0.4)
```
:::

::: {.column width="50%"}

- Even though Colombo, Gampaha, and Kalutara are geographically close in the Western Province,
their dengue time-series patterns can look quite different.

- Spatial proximity does not imply temporal similarity in dengue dynamics.
:::
::::


# {background-image="map7.png" background-size="contain" background-repeat="no-repeat" background-position="center"}

# There is no centralized repository providing readily accessible, analysis-ready dengue data.

# Actions
